---
title: "Practical homework 2"
author: "Devlin Hoang"
date: "2026-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(gbm)
library(randomForest)
library(e1071)
library(caret)
library(tidyverse)
library(scales)
library(gridExtra)
library(pROC)
set.seed(42)
#Read data and remove NA
health<- read.csv('nhis_2022.csv', stringsAsFactors = FALSE)
cat('Helth dimension:', nrow(health), 'row x', ncol(health), 'cols\n')
```

```{r}
df <- health %>%
  filter(AGE>=18, AGE<997)%>%
  filter(DIABETICEV %in% c(1,2))%>%
  mutate(
    diabetes= factor(
      ifelse(DIABETICEV==2, 'Yes','No'),
      levels= c('No','Yes') #'No' is the reference/negative class
    )
  )%>%
  
#Organize by sex
filter(SEX %in% c(1,2))%>%
mutate(sex_female= as.integer(SEX==2))%>%
  
#Organize by martial status:
filter(MARSTCUR %in% 1:6)%>%
mutate(partnered = as.integer(MARSTCUR %in% c(1,6)))%>%

#Organize by education
filter(EDUC %in% c(103, 116, 201, 202, 301, 302, 303, 400)) %>%
  mutate(
    educ_ord = case_when(
      EDUC %in% c(103, 116) ~ 0,   # Less than HS
      EDUC %in% c(201, 202) ~ 1,   # HS diploma/GED
      EDUC %in% c(301, 302) ~ 2,   # Some college / AA
      EDUC == 303           ~ 3,   # Bachelor's
      EDUC == 400           ~ 4    # Graduate degree
    )
  ) %>%
#Organize by income: 
mutate(
  poverty_grp= case_when(
    POVERTY %in% 11:13 ~0, #low income
    POVERTY %in% 21:31 ~1, #Middle class
    POVERTY %in% 32:37~2, #Higher income
  )
)%>%
#Rename BMICALC as BMI and store as a float with values >=90
filter(BMICALC<90)%>%
rename(bmi=BMICALC)%>%
#Convert sleep hurs and rename HRSLEEP to sleep_hr
filter(HRSLEEP>=1, HRSLEEP<=24)%>%
rename(sleep_hr=HRSLEEP)%>%
#Organize based on physical activities: 
mutate(
  mod_min  = ifelse(MOD10DMIN == 996, 0L, MOD10DMIN),
    vig_min  = ifelse(VIG10DMIN == 996, 0L, VIG10DMIN)
  ) %>%
filter(mod_min < 990, vig_min < 990) %>%
  
#Filter by fruit and vegetable consumption habits: 
#0-9 is valid, 90+ is missing (drop)
filter(FRUTNO <90, VEGENO<90)%>%
rename(fruit_wk= FRUTNO, veg_wk=VEGENO)%>%
#Filter unhealthy diet habit
filter(SODAPNO<90, PIZZANO<90, FRIESPNO<90)%>%
rename(soda_wk=SODAPNO, pizza_wk=PIZZANO, fries_wk=FRIESPNO)%>%
#Filter alcohol consumption habit: 
mutate(alc_days= ifelse(ALCANYNO==996, 0L, ALCANYNO))%>%
filter(alc_days>=0, alc_days<=365)%>%
#Filter based on working hours:
filter(HOURSWRK<97)%>%
rename(work_hrs=HOURSWRK)

cat('After cleaning data:', nrow(df),'adults with complete records\n\n')
cat('Diabetes prevalance: \n')
print(prop.table(table(df$diabetes)))
```
Clean up and preparing data for modeling

```{r}
#Final predictor set:
#Demographics: age, sex, partnered, education, poverty
#Boday/health: bmi, sleep hours
#Physical act: mod_min, vig_min
#Diet: fruit, vegetable, fries, pizza, soda
#Lifestyle: alcohol per day, work hours
model_cols<- c(
  'diabetes',
  'AGE','sex_female','partnered','educ_ord','poverty_grp',
  'bmi','sleep_hr',
  'mod_min','vig_min',
  'fruit_wk','veg_wk','soda_wk','pizza_wk','fries_wk',
  'alc_days','work_hrs')
df_model <- df %>%
  select(all_of(model_cols))%>%
  drop_na()
cat('Model-ready dataset: ', nrow(df_model),'rows', ncol(df_model)-1,'predictors \n')
```
```{r}
#Exploratory data analysis: 
#Age distribution by diabetes status:
p_age<- ggplot(df_model, aes(x=AGE, fill= diabetes)) + geom_density(alpha=0.5) + scale_fill_manual(values= c('No'='steelblue', 'Yes'='red')) + labs(title= 'Age Distribution by Diabetes Status',
                                                                                                                                                     x= 'Age (years)', y='Density', fill='Diabetes') + theme_minimal(base_size=12)

#BMI by diabetes status: 
p_bmi<- ggplot(df_model, aes(x= diabetes, y=bmi, fill= diabetes)) + geom_violin(alpha=0.6, trim= TRUE) + geom_boxplot(width=0.15, alpha=0.8, outlier.size=0.3) + scale_fill_manual(values= c('No'= 'skyblue','Yes'='darkred')) + labs(title= 'BMI by Diabetes Status', x='Diabetes', y='BMI') + theme_minimal(base_size=12) + theme(legend.position='none')

#Vigorous activity:
p_vig<- ggplot(df_model, aes(x= diabetes, y=vig_min, fill=diabetes))+ geom_boxplot(alpha=0.7, outlier.size=0.3) + scale_fill_manual(values= c('No'='blue', 'Yes'= 'tan')) + labs(title='Vigorous Activity (min) by Diabetes', x='Diabetes', y='Minutes vigorous activity/day')+ theme_minimal(base_size=12)+ theme(legend.position='none')

#Sleepp hour:
p_sleep<- ggplot(df_model, aes(x=sleep_hr, fill=diabetes))+ geom_histogram(position= 'identity',alpha=0.5,binwidth=1) + scale_fill_manual(values= c('No'='darkblue', 'Yes'= 'tomato')) + labs(title='Sleep hours per Night by Diabetes', x='Hours of sleep', y='Count', fill='Diabetes')+ theme_minimal(base_size=12)

#Education vs diabetes prevalance: 
p_educ <- df_model %>%
  group_by(educ_ord, diabetes) %>% summarise(n = n(), .groups = "drop") %>%
  group_by(educ_ord) %>% mutate(pct = n / sum(n)) %>%
  filter(diabetes == "Yes") %>%
  ungroup()%>%
  mutate(educ_label = factor(educ_ord,
    labels = c("< HS","HS/GED","Some college","Bachelor's","Graduate"))) %>%
  ggplot(aes(x = educ_label, y = pct)) +
  geom_col(fill = "salmon", alpha = 0.8) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Diabetes Prevalence by Education", x = NULL, y = "% with diabetes") +
  theme_minimal(base_size = 12)

#Soda consumption vs diabetes: 
p_soda<- df_model %>%
  mutate(soda_grp= cut(soda_wk, breaks=c(-1,0,1,3,9),
                       labels= c('0','1','2-3','4+')))%>%
  group_by(soda_grp, diabetes) %>% summarise(n=n(), .groups='drop')%>%
  group_by(soda_grp) %>% mutate(pct=n/sum(n)) %>%
  filter(diabetes=='Yes') %>%
  ggplot(aes(x=soda_grp, y=pct))+ geom_col(fill='brown', alpha=0.8) + scale_y_continuous(labels= percent_format())+ labs(title= ' Diabetes Prevalance by Soda/Week', x='Soda per week', y='% with diabetes') + theme_minimal(base_size=12)

eda_grid <- grid.arrange(p_age, p_bmi, p_vig, p_sleep, p_educ, p_soda, ncol=2)
```

Create different plots for different predictors vs diabetes

```{r}
#Train/test split
n_yes<- sum(df_model$diabetes=='Yes')
n_no<- sum(df_model$diabetes=='No')
cat('Raw class count -- Yes:',n_yes,'| No: ',n_no, '\n')

df_yes<- df_model %>% filter(diabetes=='Yes')
df_no_sub<- df_model %>% filter(diabetes=='No') %>% sample_n(n_yes)
df_balance<- bind_rows(df_yes,df_no_sub)%>%
  slice_sample(prop=1)
cat('Balance dataset: ',nrow(df_balance),'|Yes:',sum(df_balance$diabetes=='Yes'),'|No:',sum(df_balance$diabetes=='No'))

#75/25 train test split:
train_idx<- createDataPartition(df_balance$diabetes, p=0.75, list=FALSE)
train_df<- df_balance[train_idx,]
test_df<- df_balance[-train_idx,]
cat('Train: ', nrow(train_df), '|Test: ', nrow(test_df),'\n')

#Standardize continuous predictors (mu=0, sigma=1)
continuous_cols<- c('AGE','bmi','sleep_hr','mod_min','vig_min','fruit_wk','veg_wk','soda_wk','pizza_wk','fries_wk','alc_days','work_hrs')
preproc<- preProcess(train_df[,continuous_cols],method= c('center','scale'))
train_scaled<- predict(preproc, train_df)
test_scaled<- predict(preproc, test_df)
```
Train and test split for dataset with 75/25. Also downsaple for majority class predictors, some predictors were kept as is due to how narrow they are on the integer scale. 
```{r}
#Create function to evaluate SVM;
evaluate_svm<- function(model, test_data, label){
  preds<- predict(model, newdata= test_data)
  cm<- confusionMatrix(preds,test_data$diabetes, positive='Yes')
  roc_obj<- roc(as.numeric(test_data$diabetes =='Yes'),
                as.numeric(preds=='Yes'), quiet=TRUE)
  cat('\n======',label,'======\n')
  cat(sprintf('Accuracy: %.4f\n', cm$overall['Accuracy']))
  cat(sprintf('Sensitivity: %.4f\n', cm$byClass['Sensitivity']))
  cat(sprintf('Specificity: %.4f\n', cm$byClass['Specificity']))
  cat(sprintf('Precision: %.4f\n', cm$byClass['Precision']))
  cat(sprintf('F1 score: %.4f\n', cm$byClass['F1']))
  cat(sprintf('AUC: %.4f\n', auc(roc_obj)))
  cat('\nConfusion Matrix:\n')
  print(cm$table)
  
  list(label = label,
       accuracy= unname(cm$overall['Accuracy']),
       sens= unname(cm$byClass['Sensitivity']),
       spec= unname(cm$byClass['Specificity']),
       prec= unname(cm$byClass['Precision']),
       f1= unname(cm$byClass['F1']),
       auc= as.numeric(auc(roc_obj)),
       roc= roc_obj,
       cm= cm,
       model= model)
}
```

```{r}
#Fit SVM  linear models: 
tune_linear<- tune(
  svm, 
  diabetes~.,
  data= train_scaled,
  kernel='linear',
  ranges= list(cost=c(0.001,0.01,0.1,1,10)),
  tunecontrol=tune.control(cross=5)
)
cat('Best C: ', tune_linear$best.parameters$cost,'\n')
cat('CV error: ',tune_linear$best.performance,'\n')
res_linear<-evaluate_svm(tune_linear$best.model, test_scaled, 'Linear svm')

#Fit SVM radial models: 
tune_radial<- tune(
  svm, 
  diabetes~.,
  data= train_scaled,
  kernel='radial',
  ranges= list(cost=c(0.1,1,10,100),
               gamma= c(0.001,0.01,0.1,0.5)),
  tunecontrol=tune.control(cross=5)
)
cat('Best C: ', tune_radial$best.parameters$cost,'|Best gamma:',tune_radial$best.parameters$gamma,'\n')
cat('CV error: ', tune_radial$best.performance,'\n')
res_radial<-evaluate_svm(tune_radial$best.model, test_scaled, 'Radial svm')

#Fit SVM  polynomial models: 
tune_polynomial<- tune(
  svm, 
  diabetes~.,
  data= train_scaled,
  kernel='polynomial',
  ranges= list(cost=c(0.1,1,10),
               degree= c(2,3),
               coef0= c(0,1)),
  tunecontrol=tune.control(cross=5)
)
cat('Best C: ', tune_polynomial$best.parameters$cost,'|Best degree:',tune_polynomial$best.parameters$degree,'|Best coefficient:',tune_polynomial$best.parameters$coef0,'\n')
cat('CV error: ',tune_polynomial$best.performance,'\n')
res_poly<-evaluate_svm(tune_polynomial$best.model, test_scaled, 'Polynomial svm')


```

Create and fit the dataset with different type of kernels

```{r}
#Model comparison: 
comparison<- tibble(
  Kernel= c('Linear','Radial','Polynomial'),
  Accuracy= c(res_linear$accuracy, res_radial$accuracy, res_poly$accuracy),
  sensitivity= c(res_linear$sens, res_radial$sens, res_poly$sens),
  specificity= c(res_linear$spec, res_radial$spec, res_poly$spec),
  F1= c(res_linear$f1, res_radial$f1, res_poly$f1),
  AUC= c(res_linear$auc, res_radial$auc, res_poly$auc)
)%>% mutate(across(where(is.numeric),~round(.,4)))

cat('\n\n===== Model Comparison Table ===\n')
print(comparison)
```
Create comparing table for all kernels type, with Radial seems to perform slightly bettern than linear.

```{r}
roc_df<- bind_rows( tibble(fpr=1- res_linear$roc$specificities, 
                           tpr= res_linear$roc$sensitivities, kernel='Linear'), 
                    tibble(fpr=1 - res_radial$roc$specificities, 
                           tpr= res_radial$roc$sensitivities, kernel='Radial'), 
                    tibble(fpr=1 - res_poly$roc$specificities, 
                           tpr= res_poly$roc$sensitivities, kernel='Polynomial'))
roc_plot <- ggplot(roc_df, aes(x = fpr, y = tpr, color = kernel)) +
  geom_line(linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(
    values = c('Linear' = 'steelblue',
               'Radial' = 'salmon',
               'Polynomial' = 'darkgreen'),
    labels = c(
      "Linear" = paste0("Linear (AUC=", round(res_linear$auc, 3), ")"),
      "Radial" = paste0("Radial (AUC=", round(res_radial$auc, 3), ")"),
      "Polynomial" = paste0("Polynomial (AUC=", round(res_poly$auc, 3), ")")
    )
  ) +
  labs(
    title = "ROC Curves -- SVM Kernel Comparison",
    subtitle = "Target: Diabetes (DIABETICEV) | NHIS 2022",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)",
    color = "Kernel"
  ) +
  coord_equal() +
  theme_minimal(base_size = 13) +
  theme(legend.position = c(0.7, 0.2))
print(roc_plot)
```

Generate ROC curve for all SVM models 

```{r}
#Analysis based on SEX: 
best_models<- list(res_linear, res_radial, res_poly)
best_idx<- which.max(c(res_linear$auc, res_radial$auc, res_poly$auc))
best_res<- best_models[[best_idx]]
cat("\n\n=== SUBGROUP ANALYSIS BY SEX (best model:", best_res$label, ") ===\n")

for(sex_val in c(0,1)){
  label<- ifelse(sex_val==1, 'Female','Male')
  subset<- test_scaled %>% filter(sex_female==sex_val)
  if(nrow(subset)<20)next
  preds<- predict(best_res$model, newdata=subset)
  cm<- confusionMatrix(preds, subset$diabetes, positive='Yes')
  roc_s<- roc(as.numeric(subset$diabetes=='Yes'),
              as.numeric(preds=='Yes'),quiet=TRUE)
  cat(sprintf('\n[%s | n=%d | diabetes=%.1f%%]\n',
      label,nrow(subset),mean(subset$diabetes=='Yes')*100))
  cat(sprintf('Accuracy: %.4f| F1: %.4f |AUC: %.4f\n',
              cm$overall['Accuracy'], cm$byClass['F1'],auc(roc_s)))
}

#Analysis subgroup based on age (50+):
cat('\n\n==== Group Analysis by AGE(50+)=====\n')
test_50<- test_df%>% filter(AGE>=50)
test_50_sc<-predict(preproc, test_50)

cat('Test observation age 50+: ',nrow(test_50_sc),'\n')
pred_50<- predict(best_res$model, newdata= test_50_sc)
cm50<- confusionMatrix(pred_50, test_50_sc$diabetes, positive='Yes')
roc_50<- roc(as.numeric(test_50_sc$diabetes =='Yes'),
             as.numeric(pred_50 =='Yes'), quiet=TRUE)
 cat(sprintf('Accuracy: %.4f| F1: %.4f |AUC: %.4f\n',
              cm$overall['Accuracy'], cm$byClass['F1'],auc(roc_s)))
``` 

